Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-15 paper/paper.tex
"Liveness analysis - clarified the difference in LA of eager and lazy"

To describe  the issues  in formulating liveness  of a  lazy language,
consider the function definition shown::

\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} (\DEFINE\ (\pf~\pa)
	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
	\end{uprogram}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The main challenges in doing liveness analysis for lazy languages are, 
1) a variable can either point to a fully evaluated data value or it 
could point to a closure which on evaluation generates the required data value. For example, 
consider the statement {\tt let \py~ $\leftarrow $ (\CDR~\px)}. In case of an eager language 
the variable \py~ points to  the \CDR~ part of \px~ which is guaranteed to be a cons cell. 
Hence any demand on \py~ can be immediately transferred to the \CDR~ part of \px. In case of 
a lazy language, \px~ might not be evaluated and  hence pointing to a closure. In our 
example, it points to the closure corresponding to (\CONS~1~\px). It does not make any sense 
to transfer the demand onto the \CDR~ part of \px~ as it still does not exist.
, 2) it is not possible to statically determine when a closure gets evaluated.  Therefore 
during GC we treat a closure as a demand transformer and tranform the demand on to the arguments 
of the closure pointed. This is similar to the process of transferring the demand 
from the return value of a function to the arguments of the function call in an eager language.

In our example, during GC if we find that \px~ points to a closure instead of a cons cell we 
find out the demand/liveness on the arguments of the closure, namely \py.
Statically it cannot be determined what values a closure points to or when it gets evaluated,
hence our method approximates the liveness of closure arguments at runtime 
by tagging extra information with each closure. Since the arguments of the closures could themselves be closures in turn, 
we recursively apply the same procedure until we hit a data value. The only case where we do not 
follow this procedure is if the demand on a particular closure is null. In such a case we safely collect 
the entire closure as garbage.

We now describe the process of finding out the liveness 
of each variable defined in the function $\pf$ assuming a symbolic demand $\sigma$ on it. 
For obvious reasons, this is also the demand on the return values $u$ and $y$, and thus the
liveness  at  the  program  points  $\pi_7$ and  $\pi_8$  are  $\lbrace
\pu.\sigma \rbrace$ and $\{\py.\sigma\}$.  The liveness at $\pi_7$ and
$\pi_8$ are propagated backwards as in traditional liveness analysis.

@@

since  the liveness  of  \pu\ will  not  be consulted  by the  garbage
collector at $\pi_6$.

@@

The liveness of lazy and eager evaluation differ in two respects:
\begin{enumerate}
\item  In an  eager  language, \pw\  would  not have  been live  after

@@

\item The second difference is that  our \LET s can be recursive as in
  program point $\pi_1$.  In such a case, the liveness of a let has to
  be formulated as a fix-point computation.
\end{enumerate}

  To address the first issue, we  regard the body of the function as a
  tree and consider paths from the root of this tree to program points
  which     trigger     a     evaluation---\SIF\    conditions     and
  \RETURN\  expressions. We call  such paths  {\em E-paths}.   For the
  program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
  \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
  analysis consists of two maps:  The map $\cal{M}$ maps a return path
  to a liveness  environment that is common to  all the program points
  in the path. And a second  map $\cal{P}$ maps a program point to the
  set of  return paths on  which the point  lies.  If a  program point
  happens  to fall  on  more than  one  such path,  then the  liveness
  environment of the  program point is the variable-wise  union of the
  liveness  environments  of the  individual  paths.   In the  example
  program,  ${\cal  M}(EP_1)=  \ldots$,  ${\cal  M}(EP_2)=\ldots$  and
  ${\cal M}(EP_3)= \ldots$.  Since  ${\cal P}(pi_4)$ is $\lbrace EP_1,
  EP_2, EP_3 \rbrace$, the liveness of $pi_4$ is \ldots.

>>>>

To describe  the issues  in formulating liveness  of a  lazy language,
consider the function definition shown:

\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} (\DEFINE\ (\pf~\pa)
	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
	\end{uprogram}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\medskip

Liveness analysis for lazy languages differ from eager languages in
two respects:
\begin{enumerate}
\item Unlike eager languages, a variable may be  bound to an
  expression that is not fully evaluated. i.e. a closure.
\item The point where an expression gets evaluated cannot be statically
  determined.
\end{enumerate}

To  deal  with the  first  issue, we  apply  the  results of  liveness
analysis only  to fully evaluated  data.  If the marking  phase during
garbage collection leads to a closure, and the closure is not dead, we
mark the entire reachable memory  under the closure for collection.  A
better scheme  would be  to garbage collect  the closure based  on the
liveness  at the point  of its  creation.  This  is sound  because our
language is statically scoped.  However, this is work in progress, and
the method described in this paper switches over to reachability while
collecting closures.

%% However, we GC,  if we find a variable can either  point to a fully
%% evaluated  data value  or  it could  point  to a  closure which  on
%% evaluation  generates  the   required  data  value.   For  example,
%% consider the statement {\tt  let \py~ $\leftarrow $ (\CDR~\px)}. In
%% case of  an eager  language the variable  \py~ points to  the \CDR~
%% part of  \px~ which  is guaranteed  to be a  cons cell.   Hence any
%% demand on \py~ can be  immediately transferred to the \CDR~ part of
%% \px.  In case  of a lazy language, \px~ might  not be evaluated and
%% hence  pointing to  a closure.  In our  example, it  points  to the
%% closure corresponding to (\CONS~1~\px).  It does not make any sense
%% to transfer the demand onto the \CDR~ part of \px~ as it still does
%% not exist.  , 2) it is  not possible to statically determine when a
%% closure gets evaluated.  Therefore during  GC we treat a closure as
%% a demand transformer and tranform the demand on to the arguments of
%% the  closure   pointed.   This  is   similar  to  the   process  of
%% transferring the demand from the  return value of a function to the
%% arguments of the function call in an eager language.

%% In our example, during GC if  we find that \px~ points to a closure
%% instead  of a  cons cell  we find  out the  demand/liveness  on the
%% arguments  of the  closure, namely  \py.  Statically  it  cannot be
%% determined  what  values  a  closure  points to  or  when  it  gets
%% evaluated, hence  our method  approximates the liveness  of closure
%% arguments  at  runtime  by  tagging  extra  information  with  each
%% closure.  Since  the arguments of the closures  could themselves be
%% closures in turn, we recursively  apply the same procedure until we
%% hit  a data  value.  The only  case  where we  do  not follow  this
%% procedure is if the demand on a particular closure is null. In such
%% a case we safely collect the entire closure as garbage.

To see  our way of  handling the second  issue, we trace  the liveness
analysis to  an extent. Assume  that the calls  to the $\pf$  place an
overall demand  of $\sigma$ on the  body of \pf.  on  the now describe
the process  of finding out the  liveness of each  variable defined in
the function  $\pf$ assuming  a symbolic demand  $\sigma$ on  it.  For
obvious reasons, this is also the  demand on the return values $u$ and
$y$, and thus  the liveness at the program  points $\pi_7$ and $\pi_8$
are $\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The liveness
at  $\pi_7$ and  $\pi_8$ are  propagated backwards  as  in traditional
liveness analysis.

@@

since  \pu\  is yet  to  come into  existence at  $\pi_6$ and  its
liveness will not be consulted by the garbage collector at this point.

@@

In an  eager  language, \pw\  would  not have  been live  after

@@

%% The second  difference is that  our \LET s  can be recursive  as in
%%   program point $\pi_1$.  In such a case, the liveness of a let has
%%   to be formulated as a fix-point computation.
  To address this issue, we  regard the body of the function as a
  tree and consider paths from the root of this tree to program points
  which     trigger     a     evaluation---\SIF\    conditions     and
  \RETURN\  expressions. We call  such paths  {\em E-paths}.   For the
  program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
  \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
  analysis consists of two maps:  The map $\cal{M}$ maps a return path
  to a liveness  environment that is common to  all the program points
  in the path. And a second  map $\cal{P}$ maps a program point to the
  set of  return paths on  which the point  lies.  If a  program point
  happens  to fall  on  more than  one  such path,  then the  liveness
  environment of the  program point is the variable-wise  union of the
  liveness  environments  of the  individual  paths.   In the  example
  program,  ${\cal  M}(EP_1)=  \ldots$,  ${\cal  M}(EP_2)=\ldots$  and
  ${\cal M}(EP_3)= \ldots$.  Since  ${\cal P}(pi_4)$ is $\lbrace EP_1,
  EP_2, EP_3 \rbrace$, the liveness of $pi_4$ is \ldots.
